Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-17 paper/paper.tex
update

\begin{abstract}
We introduce a method that uses generic programming tensor network to compute various properties of a class of hard graph problems that can be decribed by local constraints.
Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph.
Other hard graph problems solvable in our framework includes the cutting problem, the vertex coloring problem, the maximal clique problem, the dominating set problem, and satisfiability problem, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
We show how our method helps people understand these hard problems better.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there is an important class of problems that can be local constraints over a vertex and its neighborhood,
including the independent set problem, the cutting problem, the vertex coloring problem, the satisfiability problem,
the maximal clique problem, and the vertex cover problem et al ~\cite{Moore2011}.
%There is a widespread interest of these problem because they have wide range of applications in scheduling, logistics,
%wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
%\blue{need more diversed applications not limited to IS}
Many decision and counting properties of these problems are in general intractable at large sizes (exponential runetime in problem size in the worst case),
but it is nevertheless desirable to have high-performance algorithms because of their ubiquitous applications~\cite{Butenko2003, Wu2015}.

The most studied version of the above problems is finding a maximum/minimum set satisfying the contraints and the corresponding set size,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. it is unlikely to solve them in a polynomial time.
\purple{Is MIS really the most studied NP-hard optimization problem? I would guess it is heavily studied, but not necessarily the most studied.}
\blue{Here, we haven't mentioned the independent set, here ``version'' means, e.g. counting, enumeration, devision. I think finding one of the largest/smallest set and its cardinality in the above mentioned problems is the most studied?}
In this paper, we focus on the less studied \textit{solution space property} computation which can be harder, e.g. counting independent sets of a general graph is in \#P-complete~\cite{Vadhan2001}, but is crutial towards understanding a hard problem.
Here, the \textit{solution space property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version of the above problems, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
Solution space property computation allows us to understand the solution space better.
With the counting of sets at different sizes, one can related the computational hardness with the counting of sets at different sizes~\cite{Xu2018}.
With the configuration enumeration/sampling at a given size, one can measure the overlap gap property better,
while the configuration space connectivity also privides us an intuition about designing a better classical or quantum algortihms.
\purple{I would give examples of why we want to compute solution space
properties.
For instance, in statistical physics, computing thermal states and
observables via the partition function. In complexity theory, proving limits on local
algorithms (OGP).}\blue{now?}

We propose to view the computation of all these properties from an abstract algebra perspective and use a unified framework,
generic tensor-network, to solve these seeming unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network in probabilistic modeling~\cite{Bishop2006} or einsum in linear algebra libraries such as numpy.
It is composed of a set of tensors and a set of labels for specifying the contraction of tensors,
where the tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
Given a good contraction order, the space required to contract a tensor network is 2 power the treewidth of its line graph~\cite{Markov2008}.
\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time.
\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
When using tensor network for solving one of the best sets with the maximum size,
this time complexity is the same as the dynamic programming algorithm~\cite{Courcelle1990, Fomin2013}.
These two algorithms are related but are not equivalent.
Dynamic programming is more general purposed, while a tensor network has a richer algebraic structure, and this structure is crutial for solution space property computation.
\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

To avoid the discussion being unnecessarily diversed, we mainly focus on the independent set problem,
and discuss other combinatorial optimization problems,
including the cutting problem, the matching problem, the vertex coloring problem, the satisfiability problem, the dominating set problem, the set packing problem and the maximal clique problem, in \App{sec:otherproblems}.
\xpurple{More details on which combinatorial optimization problems this
applies to?}
The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce an independent set problem to a tensor network contraction problem \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting},\Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor\xpurple{signifying which pairs of indices are summed over}.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power,
while the generalized tensor network might produce smaller contraction complexity,
which will be illustrated in \App{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.
\xpurple{I feel that the $\delta$ tensors point is unnecessary to mention in the main text. Because not much detail is given, I internalized what it meant only on the third read-through.}

@@

\purple{I think the following paragraph could be clarified. When you read the definition
of generic programming, it's not clear what ``in the most general setting'' means. My
understanding is it just means that the program will work regardless of the input data
type - it would be good to say this explicitly. }
\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
\purple{I also think the way it is written takes a while
to understand what dynamically versus statically typed languages are. The definition,
“When these languages “see” a new input type, the compiler can recompile the generic
program for the new type.” comes rather late in the paragraph.}
\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
lost on someone only with a physics background.}
This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
To understand the first aspect on generality, suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about performance. For dynamically typed languages such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

This motivates us to think about what is the most general tensor element type allowed in a tensor network contraction program.
We find that as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we wish this contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
The Table~\ref{tbl:generictypes} summarizes those properties that can be computed by various tensor element types.

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their numbers
   \purple{``and their counting'' is gramatically confusing, can we say independence polynomial or degeneracy or something?}\blue{Since you mentioned degeneracy is bad, what about ``numbers''?}} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Sum Product Tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with Bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with Set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

It is in general not computational efficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products, which scales exponentially as the number of vertices.
\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a tensor network is to contract two tensors at a time with a certain order utilizing the associativity and commutativity of tensor contraction.
A fully optimized contraction order can reduce the time complexity significantly, while requiring a minimum space complexity $O(2^{\text{tw}(G)})$~\cite{Markov2008} to store intermediate contraction results.
\purple{Perhaps just say time and space
complexity? Order of magnitude doesn't make sense to me in terms of scaling. Also,
what is the time complexity of finding a good contraction order? Is there a known
average-case complexity? This seems important.}\blue{see added texts below.}
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
In practise, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth itself is NP-hard.
However, it is easy to find a close to optimal contraction order in typically a few minutes with a heuristic contraction order finding algorithm~\cite{Kourtis2019,Kalachev2021}.
For large scale applications,
it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
enumerating certain degrees of freedom in an external loop so that the space complexity of contracting a smaller tensor network inside the loop can be smaller.

\begin{example}\label{eg:tensorcontraction}
In the following, we mapping a 5 vertex graph (left) to a tensor network (right) and show how the contraction order simplifies the computation.

@@

\begin{equation}
\eqname{S1}
\begin{split}
    \sigma \oplus \tau &= {\rm select}(\sigma, \tau), \\
    \sigma \odot \tau &= (\sigma\lor^\circ \tau),\\
    \mymathbb{0} &= 1^{\otimes |V|}, \\
    \mymathbb{1} &= 0^{\otimes |V|},
\end{split}\label{eq:singleconfig}
\end{equation}
where the \texttt{select} function picks one of $\sigma$ and $\tau$ by some criteria to make the algebra commutative and associative, e.g. by picking the one with a smaller integer value.

\subsection{Sampling extremely large configuration space}
When the graph size goes large, the configurations might be impossible to fit into any type of storage.
Then instead of using the set algebra, one can use a binary sum-product tree as a compact representation of the configurations.
We use a quadruple $(type, data, left, right)$ to represent a node in the tree, where $type$ is one of
\texttt{LEAF}, \texttt{ZERO}, \texttt{SUM} and \texttt{PROD}, $data$ is a bit string as the content in a \texttt{LEAF} node,
$left$ and $right$ are left and right operands of \texttt{SUM} and \texttt{PROD} nodes.

\begin{equation}
\eqname{SPT}
\begin{split}
    s \oplus t &= (\texttt{SUM}, /, s, t)\\
    s \odot t &= (\texttt{PROD}, /, s, t)\\
    \mymathbb{0} &= (\texttt{ZERO}, /, /, /)\\
    \mymathbb{1} &= (\texttt{LEAF}, 0^{\otimes |V|}, /, /).
\end{split}\label{eq:expr}
\end{equation}
\purple{It might be nice to have a picture of an example tree if you have time. However,
it is pretty clear from the definition, it would just help to visualize. Do you ever use the
zero element?}\blue{now we have an example, but it does not include a zero element, do you think I should use a more concrete one instead of an abstract one?}
The vertex tensor and edge tensor become
\begin{equation}
    W^{\rm SPT}((\texttt{LEAF},\boldsymbol{e}_i,/,/)) = \left(\begin{matrix}
        \mymathbb{1} \\
        (\texttt{LEAF},\boldsymbol{e}_i,/,/)
    \end{matrix}\right),   
    \qquad
        B^{\rm SPT} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

>>>>

\begin{abstract}
We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph.
Other hard graph problems solvable in our framework includes the cutting problem, the vertex coloring problem, the maximal clique problem, the dominating set problem and the satisfiability problem, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
We show how our method helps to understand these hard problems better.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there is an important class of problems that can be local constraints over a vertex and its neighborhood,
including the independent set problem, the cutting problem, the vertex coloring problem, the satisfiability problem,
the maximal clique problem, and the vertex cover problem~\cite{Moore2011}.
%There is a widespread interest of these problem because they have wide range of applications in scheduling, logistics,
%wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
%\blue{need more diversed applications not limited to IS}
Many decision and counting properties of these problems are in general intractable at large sizes (exponential runetime in problem size in the worst case),
but it is nevertheless desirable to have high-performance algorithms because of their ubiquitous applications~\cite{Butenko2003, Wu2015}.

The most studied version of the above problems is finding a maximum/minimum set satisfying the contraints and the corresponding set size,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. it is unlikely to be solved in a polynomial time.
\purple{Is MIS really the most studied NP-hard optimization problem? I would guess it is heavily studied, but not necessarily the most studied.}
\blue{Here, we haven't mentioned the independent set, here ``version'' means, e.g. counting, enumeration, devision. I think finding one of the largest/smallest set and its cardinality in the above mentioned problems is the most studied?}
In this paper, we focus on the less studied \textit{solution space property} computation which can be harder, e.g. counting independent sets of a general graph is in \#P-complete~\cite{Vadhan2001}, but is crucial towards understanding a hard problem.
Here, the \textit{solution space property} refers to a class of quantities that not only include the maximum/minimum set size,
but also include the number of sets at a give size, enumeration of all sets at a give size or direct sampling of such sets when they are not enumeratable.
For the weighted version of the above problems, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.
Solution space property computation allows us to understand the solution space better.
With the counting of sets at different sizes, one can related the computational hardness with the counting of sets at different sizes~\cite{Xu2018}.
With the configuration enumeration/sampling at a given size, one can measure the overlap gap property better,
while the configuration space connectivity also privides us an intuition about designing a better classical or quantum algortihms.
\purple{I would give examples of why we want to compute solution space
properties.
For instance, in statistical physics, computing thermal states and
observables via the partition function. In complexity theory, proving limits on local
algorithms (OGP).}\blue{now?}

We propose to view the computation of all these properties from an abstract algebra perspective and use the unified framework of
generic tensor-network to solve these seemingly unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network in probabilistic modeling~\cite{Bishop2006} or einsum in linear algebra libraries such as numpy.
It is composed of a set of tensors and a set of labels for specifying the contraction of tensors,
where the tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
Given a good contraction order, the space required to contract a tensor network is 2 power the treewidth of its line graph~\cite{Markov2008}.
\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time.
\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
When using tensor network for solving one of the best sets with the maximum size,
this time complexity is the same as the dynamic programming algorithm~\cite{Courcelle1990, Fomin2013}.
These two algorithms are related but are not equivalent.
Dynamic programming is more general purposed, while a tensor network has a richer algebraic structure, and this structure is crucial for solution space property computation.
\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

To avoid the discussion being unnecessarily diversed, we mainly focus on the independent set problem,
and discuss other combinatorial optimization problems,
including the cutting problem, the matching problem, the vertex coloring problem, the satisfiability problem, the dominating set problem, the set packing problem and the maximal clique problem, in \App{sec:otherproblems}.
\xpurple{More details on which combinatorial optimization problems this
applies to?}
The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce the independent set problem to a tensor network contraction problem in \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting}, \Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
A tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor\xpurple{signifying which pairs of indices are summed over}.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power,
while the generalized tensor network might produce smaller contraction complexity,
which will be illustrated in \App{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{sec:tensorbad}.
\xpurple{I feel that the $\delta$ tensors point is unnecessary to mention in the main text. Because not much detail is given, I internalized what it meant only on the third read-through.}

@@

\purple{I think the following paragraph could be clarified. When you read the definition
of generic programming, it's not clear what ``in the most general setting'' means. My
understanding is it just means that the program will work regardless of the input data
type - it would be good to say this explicitly. }
\blue{This definition is from the book I cited, I also do not like ``in the most general setting'', maybe we just use our version?}
\purple{I also think the way it is written takes a while
to understand what dynamically versus statically typed languages are. The definition,
“When these languages “see” a new input type, the compiler can recompile the generic
program for the new type.” comes rather late in the paragraph.}
\xpurple{Also, while I know what you mean, concepts like “inlining immutable elements” and “cache miss rate” would be
lost on someone only with a physics background.}
This definition of generic programming contains two major aspects: a single program works in the most general setting and efficiency.
To understand the first aspect on generality, suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $O \left( \log(n) \right)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about performance. For dynamically typed languages, such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as optimize the memory layout of immutable elements with fixed sizes in an array to speed up the array indexing.
In Julia, arrays with immutable elements with fixed sizes can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

This motivates us to think about what is the most general tensor element type allowed in a tensor network contraction program.
We find that, as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we require that the contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
The Table~\ref{tbl:generictypes} summarizes those properties that can be computed by various tensor element types.

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Solution space property} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their numbers
   \purple{``and their counting'' is gramatically confusing, can we say independence polynomial or degeneracy or something?}\blue{Since you mentioned degeneracy is bad, what about ``numbers''?}} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Expression tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

It is in general not computationally efficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products, which scales exponentially as the number of vertices.
\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a tensor network is to contract two tensors at a time with a certain order utilizing the associativity and commutativity of tensor contraction.
A fully optimized contraction order can reduce the time complexity significantly, while requiring a minimum space complexity $O(2^{\text{tw}(G)})$~\cite{Markov2008} to store intermediate contraction results.
\purple{Perhaps just say time and space
complexity? Order of magnitude doesn't make sense to me in terms of scaling. Also,
what is the time complexity of finding a good contraction order? Is there a known
average-case complexity? This seems important.}\blue{see added texts below.}
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
In practise, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth itself is NP-hard.
However, it is easy to find a close to optimal contraction order in typically a few minutes with a heuristic contraction order finding algorithm~\cite{Kourtis2019,Kalachev2021}.
For large scale applications,
it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
enumerating certain degrees of freedom in an external loop so that the space complexity of contracting a smaller tensor network inside the loop can be smaller.

\begin{example}\label{eg:tensorcontraction}
In the following, we map a 5-vertex graph (left) to a tensor network (right) and show how the contraction order simplifies the computation.

@@

\begin{equation}
\eqname{S1}
\begin{split}
    \sigma \oplus \tau &= \texttt{select}(\sigma, \tau), \\
    \sigma \odot \tau &= (\sigma\lor^\circ \tau),\\
    \mymathbb{0} &= 1^{\otimes |V|}, \\
    \mymathbb{1} &= 0^{\otimes |V|},
\end{split}\label{eq:singleconfig}
\end{equation}
where the \texttt{select} function picks one of $\sigma$ and $\tau$ by some criteria to make the algebra commutative and associative, e.g. by picking the one with a smaller integer value.

\subsection{Sampling extremely large configuration space}
When the graph size goes large, the configurations might be impossible to fit into any type of storage.
Then instead of using the set algebra, one can use a binary expression tree as a compact representation of the configurations.
We use a quadruple $(type, data, left, right)$ to represent a node in the tree, where $type$ is one of
\texttt{LEAF}, \texttt{ZERO}, \texttt{SUM} and \texttt{PROD}, $data$ is a bit string as the content in a \texttt{LEAF} node,
$left$ and $right$ are left and right operands of \texttt{SUM} and \texttt{PROD} nodes.

\begin{equation}
\eqname{EXPR}
\begin{split}
    s \oplus t &= (\texttt{SUM}, /, s, t)\\
    s \odot t &= (\texttt{PROD}, /, s, t)\\
    \mymathbb{0} &= (\texttt{ZERO}, /, /, /)\\
    \mymathbb{1} &= (\texttt{LEAF}, 0^{\otimes |V|}, /, /).
\end{split}\label{eq:expr}
\end{equation}
\purple{It might be nice to have a picture of an example tree if you have time. However,
it is pretty clear from the definition, it would just help to visualize. Do you ever use the
zero element?}\blue{now we have an example, but it does not include a zero element, do you think I should use a more concrete one instead of an abstract one?}
The vertex tensor and edge tensor become
\begin{equation}
    W^{\rm EXPR}((\texttt{LEAF},\boldsymbol{e}_i,/,/)) = \left(\begin{matrix}
        \mymathbb{1} \\
        (\texttt{LEAF},\boldsymbol{e}_i,/,/)
    \end{matrix}\right),   
    \qquad
        B^{\rm EXPR} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
